Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-25 paper.tex
Updates from Overleaf

Inferring evolutionary trees while accounting for both mutation and
recombination is a long-standing computational challenge. This challenge is
particularly acute for SARS-CoV-2, given the large amount of data that has been
shared, with over 15 million sequences in the GISAID database
\citep{Shu2017-hp}. It is well known that recombination distorts phylogenies
\citep{Schierup2000-fg} and affects the results of downstream analysis, e.g.,
inference of selection~\citep{Anisimova2003-vr}. However, standard phylogenetic
methods do not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Alternatively,
genetic ancestries of recombining lineages can be represented by ancestral
recombination graphs (ARGs)~\citep{Griffiths1998-ancestral,Wong2023-efficient},
which
capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.

Approaches to infer ARGs now exist that can scale to tens of thousands of human genomes and beyond  \citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which deal with levels of sequential recombination far in excess of those seen in viral phylogenies. For example, using over 90 million variable sites \citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands of modern and ancient human genomes, accounting for millions of recombination events (over 5 million "local trees"). To achieve this, \cite{Wohns2022-th} took advantage of several computational advances in ARG inference: (1) the "succinct tree sequence" data structure that can compactly store large ARGs \citep{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM) implementation of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between genomes \citep{Kelleher2019-ba}; and (3) efficient, open-source software tools \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

>>>>

%% ARGs are good for *representing* viral genealogies

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
genetic ancestries of recombining lineages can be represented by ancestral
recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if}, which
capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently, 
development of the ``succinct tree sequence'' data
structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
encoding of large ARGs, and this representation should be well-suited to storing
recombinant viral genealogies, even those with millions of sampled sequence,
such as SARS-CoV-2 (with currently over 15 million sequences in the GISAID database
\citep{Shu2017-hp}). Moreover the ``tskit'' software library, which implements this
data structure, enables efficient manipulation of the stored genealogies and is
integrated with a suite of open-source software tools to simulate, analyse, and
infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

%% ARG inference is hard

Inferring recombinant phylogenies has been a long-standing computational challenge.
However, approaches to infer ARGs now exist that can scale to tens of
thousands of human genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which
deal with levels of repeated recombination far in excess of those seen in viral
phylogenies. For example, using over 90 million variable sites
\citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands
of modern and ancient human genomes, accounting for millions of recombination
events (over 5 million "local trees"). To achieve this, \cite{Wohns2022-th}
took advantage of a highly efficient Hidden Markov Model (HMM) implementation
of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between
genomes \citep{Kelleher2019-ba}.

% What is new here?
